Fail Faster

### 2025-03-11 paper/paper.tex
a whole bunch more

\subsection{\bq}

@@

\jwc{Usual introduction to this section, corresponding to how we talked about it in the intro}

\jwc{
  \begin{itemize}
    \item To eliminate this runtime overhead and generate efficient code, we do a usual ``binding-time analysis''.
    \item The crux of this binding-time analysis is that while generators are `just' values of type \texttt{'a t}, in practice they are always \emph{completely statically defined}: one never
    computes a generator dynamically at test time.
    \item The size and the random seed are only known at run-time (the later stage), but the code of the generator itself is known at compile time.
    \item This means the generator type should be \texttt{int code -> SR.t code -> 'a code}: a statically known-function from dynamically-known int and seed to dynamically-determined result.
  \end{itemize}
}

@@

A generator of \texttt{'a}s is then a (compile-time) function \texttt{int code -> Random.t code -> 'a code}.

@@

\begin{lstlisting}
  let weighted_union (weighted_elts : (float code * 'a code) list) : 'a t =
    let sum_code = List.foldr _ in
    let go (rc : int code) (weighted_elts : (float code * 'a code) list) : 'a code =
      match weighted_elts with
      | [] -> .< failwith "Error" >.
      | (wc,xc) :: elts' ->
        .<
          if .~rc <= .~wc then .~xc
          else
            let r' = .~rc - .~wc in
            .~(go .<r'>. elts')
        >.

    fun size_c random_c ->
      .<
        let sum = .~sum_code in
        let r = SR.int .~random_c 0 sum in
        .~(go .<r>. gwcs)
      >.
\end{lstlisting}

@@

\subsection{Size Parameters, A Real Monad Instance, CPS, and ``The Trick''}

\jwc{Introduce the size parameter here. (if you deleted it beforehand.)}

\jwc{
  \begin{itemize}
    \item Careful readers might note that the types of return and bind are not actually the right types.
    \item This means that they aren't compatible with ppx-jane syntax.
    \item We could try a different type, \texttt{type 'a gen = int code -> SR.t code -> 'a}. This has an actual monad instance (show), and ou can write teh combinators of before, with \texttt{'a code gen} everywhere you had \texttt{'a gen} before.
    \item But because the result type isn't always in \texttt{code}, \texttt{bind} can't perform the let-insertion needed to preserve effect ordering.
    \item Instead, we turn to a classic technique from the multistage programming literature: writing our staged program in continuation-passing style.
    \item We define \texttt{type 'a codecps = 'z. ('a -> 'z code) -> 'z code}. This is the polymorphic CPS monad, with the continuation type always in `code'. (Footnote: this is an instance of the \emph{codensity} monad, cf ``Asymptotic Improvement of Computations over Free Monads'')
    (Cite: ``improving binding times without explicit cps conversion'', ``Multi-stage programming with functors and monads: Eliminating abstraction overhead from generic code'', and ``Closure-Free Functional Programming in a Two-Level Type Theory'')
    \item This lets us define the function \texttt{let\_insert : 'a code -> 'a codecps} as \texttt{let\_insert cx = fun k -> k .< let x = .~cx in .~(k .<x>.) >.}.
    \item of course, \texttt{codecps} has the usual monad instance for the CPS monad with a polymorphic result type.
    \item We then define \texttt{type 'a gen = int code -> SR.t code -> 'a codecps}, which lets us have it both ways. the effectful programs like \texttt{int} do their \emph{own} let-insertion, and we get a real monad instance.
  \end{itemize}
}

@@

\subsubsection{The Trick}

>>>>

@@

\jwc{Usual introduction to this section, corresponding to how we talked about it in the intro. ``We present a library that XYZ''.}

@@

Recall that staging a DSL requires carefully changing some of the types and
combinators to transform \texttt{code}s.  Deciding which types \texttt{t} can
instead be \texttt{t code} is an art known as ``binding-time analysis'' \cn:
figuring out which parts of the DSL can be determined statically (and can be
part of the compile-time stage), and which parts are only known dynamically (and
hence must be \texttt{code}).

The crux of our binding time analysis is that particular size random seed are
only known at run-time (the later stage), but the code of the generator itself
is known at compile time.  This means our library's generator type \texttt{'a
Gen.t} should be \texttt{int code -> SR.t code -> 'a code}: a compile-time
function from dynamically-known size and seed to dynamically-determined result.

@@

\begin{lstlisting}
  let weighted_union (weighted_elts : (float code * 'a code) list) : 'a t =
    let sum_code = List.foldr _ in
    let pick (rc : int code) (weighted_elts : (float code * 'a code) list) : 'a code =
      match weighted_elts with
      | [] -> .< failwith "Error" >.
      | (wc,xc) :: elts' ->
        .<
          if .~rc <= .~wc then .~xc
          else
            let r' = .~rc - .~wc in
            .~(go .<r'>. elts')
        >.

    fun size_c random_c ->
      .<
        let sum = .~sum_code in
        let r = SR.int .~random_c 0 sum in
        .~(pick .<r>. weighted_elts)
      >.
\end{lstlisting}

@@

\subsection{Size Parameters, A Real Monad Instance, CPS, and Recursion}

\subsubsection{Size Parameters}
\jwc{Introduce the size parameter here. (if you deleted it beforehand.)}

\subsubsection{CodeCPS and a Monad Instance}
\jwc{
  \begin{itemize}
    \item Careful readers might note that the types of return and bind are not actually the right types.
    \item This means that they aren't compatible with ppx-jane syntax.
    \item We could try a different type, \texttt{type 'a gen = int code -> SR.t code -> 'a}. This has an actual monad instance (show), and ou can write teh combinators of before, with \texttt{'a code gen} everywhere you had \texttt{'a gen} before.
    \item But because the result type isn't always in \texttt{code}, \texttt{bind} can't perform the let-insertion needed to preserve effect ordering.
    \item Instead, we turn to a classic technique from the multistage programming literature: writing our staged program in continuation-passing style.
    \item We define \texttt{type 'a codecps = 'z. ('a -> 'z code) -> 'z code}. This is the polymorphic CPS monad, with the continuation type always in `code'. (Footnote: this is an instance of the \emph{codensity} monad, cf ``Asymptotic Improvement of Computations over Free Monads'')
    (Cite: ``improving binding times without explicit cps conversion'', ``Multi-stage programming with functors and monads: Eliminating abstraction overhead from generic code'', and ``Closure-Free Functional Programming in a Two-Level Type Theory'')
    \item This lets us define the function \texttt{let\_insert : 'a code -> 'a codecps} as \texttt{let\_insert cx = fun k -> k .< let x = .~cx in .~(k .<x>.) >.}.
    \item of course, \texttt{codecps} has the usual monad instance for the CPS monad with a polymorphic result type.
    \item We then define \texttt{type 'a gen = int code -> SR.t code -> 'a codecps}, which lets us have it both ways. the effectful programs like \texttt{int} do their \emph{own} let-insertion, and we get a real monad instance.
  \end{itemize}
}

@@

\subsubsection{Recursive Generators}
\jwc{Describe recursion and recurisve handles}

\subsubsection{The Trick}
\jwc{Describe split --- this section is approximately experts-only, but you can just point at the Andras paper and the things it cites.}
